---
title: "v003 v085 v122 - R2025"
author: "GL"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

# about

## v003 Uninsured Adults 

### Basic information

- Description
Percentage of adults under age 65 without health insurance.

- What is the numerator?
The numerator is the number of people ages 18 to 64 who currently have no health insurance coverage. A person is uninsured if they are not currently covered by insurance through a current/former employer or union, purchased from an insurance company, Medicare, Medicaid, Medical Assistance, any kind of government-assistance plan for those with low incomes or disability, TRICARE or other military health care, Indian Health Services, VA, or any other health insurance or health coverage plan.

- What is the denominator?
The denominator is the county population ages 18 to 64.

## Raw data access

**#Website to downloaded data**
https://www.census.gov/data/datasets/time-series/demo/sahie/estimates-acs.html

**Online documentation** 
https://www.census.gov/programs-surveys/sahie/technical-documentation/methodology.html

**Directions**
Use https://www2.census.gov/programs-surveys/sahie/technical-documentation/file-layouts/sahie-file-layout-2008-2022.pdf to select the appropriate categories for:

- `agecat` (i.e., age category): = 1 (18 to 64 years)
- `racecat` (race category): = 0 (All races)
- `sexcat` (sex category): = 0 (Both sexes)
- `iprcat` (i.e., income category): =  0  (All income levels)

The measure is already calculated. But a 95% confidence interval is constructed by converting the 90% margin of error to 95%.

Confidence limits = (v003_numerator/v003_denominator) +/- ((1.96*v003_moe)/164.5)

v003_moe = pctui_moe

## v085 Uninsured

### Basic information

- Description
Percentage of population under age 65 without health insurance.

- What is the numerator?
The numerator is the number of people currently uninsured in the county under the age of 65.

- What is the denominator?
The denominator is the number of people in the county under age 65.

### Raw data access

**Website to downloaded data**
https://www.census.gov/data/datasets/time-series/demo/sahie/estimates-acs.html

**Online documentation** 
https://www.census.gov/programs-surveys/sahie/technical-documentation/methodology.html

**Directions**
Use https://www2.census.gov/programs-surveys/sahie/technical-documentation/file-layouts/sahie-file-layout-2008-2022.pdf to select the appropriate categories for:

- `agecat` (i.e., age category): = 0 (Under 65 years)
- `racecat` (race category): = 0 (All races)
- `sexcat` (sex category): = 0 (Both sexes)
- `iprcat` (i.e., income category): =  0  (All income levels)

The measure is already calculated. But a 95% confidence interval is constructed by converting the 90% margin of error to 95%.

Confidence limits = (v085_numerator/v085_denominator) +/- ((1.96*v085_moe)/164.5)

v085_moe = pctui_moe

## v122 Uninsured Children

### Basic information

- Description
Percentage of children under age 19 without health insurance.

- What is the numerator?
The numerator is the number of people under age 19 who currently have no health insurance coverage. A person is uninsured if they are not currently covered by insurance through a current/former employer or union, purchased from an insurance company, Medicare, Medicaid, Medical Assistance, any kind of government-assistance plan for those with low incomes or disability, TRICARE or other military health care, Indian Health Services, VA, or any other health insurance or health coverage plan.

- What is the denominator?
The denominator is the county population under age 19.

## Raw data access

**Website to downloaded data**
https://www.census.gov/data/datasets/time-series/demo/sahie/estimates-acs.html

**Online documentation** 
https://www.census.gov/programs-surveys/sahie/technical-documentation/methodology.html

**Directions**
Use https://www2.census.gov/programs-surveys/sahie/technical-documentation/file-layouts/sahie-file-layout-2008-2022.pdf to select the appropriate categories for:

- `agecat` (i.e., age category): = 4 (Under 19 years)
- `racecat` (race category): = 0 (All races)
- `sexcat` (sex category): = 0 (Both sexes)
- `iprcat` (i.e., income category): = 0 (All income levels)

The measure is already calculated. But a 95% confidence interval is constructed by converting the 90% margin of error to 95%.

Confidence limits = (v122_numerator/v122_denominator) +/- ((1.96*v122_moe)/164.5)

v122_moe = pctui_moe

# import packages

```{r}
library(tidyverse)
library(haven)

```

# standard county and state fips

- The final dataset will need to be aligned with standard FIPS codes and to include all counties and states.
```{r }
fips <-  bind_rows(
  read_sas("../inputs/county_fips_with_ct_old.sas7bdat") ,
  read_sas( "../inputs/state_fips.sas7bdat"))  %>% 
  arrange(statecode, countycode) %>% 
  glimpse()

```

# helpers

## get_sahie_measures()
```{r}
# func: get_sahie_measures ---------
#' get_sahie_measures(df_sahie, df_fips, age_cat, vnum)
#' This function helps calculate v003, v085 and v122 
#'
#' @param df_sahie df of sahie data set
#' @param df_fips df with standard state and county fips.
#' @param age_cat age category
#' @param vnum measure number.
#'
#' @return A dataframe.
#' @export
#'
#' @examples
#' 

get_sahie_measures <- function(df_sahie, df_fips, age_cat, vnum){
  # county + state data
  df_1 <- df_sahie %>% 
    filter(agecat == age_cat, # 1: for 18 to 64 years
           racecat == 0, sexcat == 0, iprcat == 0) %>% 
    select(-c(agecat, racecat, sexcat, iprcat))

  # national value
  df_us <- df_1 %>% 
    filter(countycode == "000") %>% 
    summarise(statecode = "00", countycode = "000",
              nipr = sum(nipr), nui = sum(nui))
  
  # county + state + us
  df_3 <- bind_rows(df_1, df_us) %>%
    arrange(statecode, countycode)
  
  # standard fips
  df_4 <- df_fips %>% 
    select(statecode, countycode) %>% 
    left_join(df_3, by = c("statecode", "countycode"))
  
  # calculate rawvalue and CI's
  df_5 <- df_4 %>% 
    select(statecode, countycode, numerator = nui, denominator = nipr, pctui_moe) %>% 
    mutate(rawvalue = numerator/denominator, .after = "denominator") %>% 
    mutate(cilow = rawvalue - 1.96 * pctui_moe/1.645/100,
           cihigh = rawvalue + 1.96 * pctui_moe/1.645/100,
           sourceflag = NA_real_) %>% 
    select(-pctui_moe)
  
  # standardize column names
  res <- df_5 %>% 
    rename(!!(paste0(vnum, "_numerator")) := numerator, 
           !!(paste0(vnum, "_denominator")) := denominator, 
           !!(paste0(vnum, "_rawvalue")) := rawvalue, 
           !!(paste0(vnum, "_cilow")) := cilow,
           !!(paste0(vnum, "_cihigh")) := cihigh,
           !!(paste0(vnum, "_sourceflag")) := sourceflag)
  
  print(dim(res))
  
  return(res)
  
}
```


## add flag_CT

- This function adds a flag to the data frame for Connecticut counties.
- 'A' = data *available* for a CT county
- 'U' = data *unavailable* for a CT county

```{r}
add_flag_CT <- function(df, vnum, old="U", new = "A"){
  CT_old_lst <- c("001", "003", "005", "007", "009", "011", "013", "015")
  CT_new_lst <- c("110", "120", "130", "140", "150", "160", "170", "180", "190")

  res <- df %>% 
    mutate(flag_CT = case_when(
      statecode == "09" & countycode %in% CT_old_lst ~ old,
      statecode == "09" & countycode %in% CT_new_lst ~ new,
      TRUE ~ NA
    )) %>% 
    rename(!!(paste0(vnum, "_flag_CT")) := flag_CT)
  
  return(res)

}

```

# raw data

```{r}
# data
sahie_raw <- read_csv("../raw_data/SAHIE/sahie-2022-csv.zip",
                       skip = 83) %>% 
  janitor::clean_names() %>% 
  glimpse()

sahie_1 <- sahie_raw %>% 
  select(statecode = statefips, countycode = countyfips, 
         agecat, racecat, sexcat, iprcat, 
         nipr, nipr_moe, nui, nui_moe, pctui_moe) %>% 
  mutate(across(c(nipr, nipr_moe, nui, nui_moe, pctui_moe), ~ as.numeric(.))) %>% 
  glimpse()
```

## variables

```{r}
tibble::tribble(
   ~var_short,  ~var_full, ~value,  ~description,
     "agecat", "Age category", 0L,  "Under 65 years",
     "agecat", "Age category", 1L,   "18 to 64 years",
     "agecat", "Age category", 2L,   "40 to 64 years",
     "agecat", "Age category", 3L,   "50 to 64 years",
     "agecat", "Age category", 4L,   "Under 19 years",
     "agecat", "Age category", 5L,   "21 to 64 years",
    "racecat", "Race category", 0L,  "All races",
    "racecat", "Race category", 1L,  "White alone",
    "racecat", "Race category", 2L,  "Black alone",
    "racecat", "Race category", 3L,   "Hispanic (any race)",
     "sexcat", "Sex category",  0L,    "Both sexes",
     "sexcat", "Sex category",  1L,     "Male",
     "sexcat", "Sex category",  2L,  "Female",
     "iprcat", "Income category", 0L, "All income levels",
     "iprcat", "Income category", 1L, "At or below 200% of poverty",
     "iprcat", "Income category", 2L, "At or below 250% of poverty",
     "iprcat", "Income category", 3L, "At or below 138% of poverty",
     "iprcat", "Income category", 4L, "At or below 400% of poverty",
     "iprcat", "Income category", 5L, "Between 138% - 400%  of poverty",
       "NIPR", "Number in demographic group for <income category>", NA, NA,
   "nipr_moe",  "MOE  for NIPR", NA,   NA,
        "NUI",  "Number uninsured", NA, NA,
    "nui_moe",  "MOE  for NUI",   NA, NA,
      "PCTUI", "Percent uninsured in demographic group for <income category>",  NA,   NA,
  "pctui_moe", "MOE  for PCTUI",  NA, NA
  ) 

```

# v003

`agecat` (i.e., age category): = 1 (18 to 64 years)
`racecat` (race category): = 0 (All races)
`sexcat` (sex category): = 0 (Both sexes)
`iprcat` (i.e., income category): =  0  (All income levels)

`nui`: numerator, Number uninsured   
`nipr`: denominator; `NIPR`: Number in demographic group for <income category>
`pctui_moe`: margin of error; `PCTUI`: Percent uninsured in demographic group for <income category>

## process data

```{r}
v003 <- get_sahie_measures(
  df_sahie = sahie_1, 
  df_fips = fips,
  age_cat = 1, # age: 18 to 64 years
  vnum = "v003") %>% 
  add_flag_CT(vnum = "v003", old="U", new = "A") %>% 
  glimpse()
```

## save data

```{r}
write_csv(v003,
          paste0("../measure_datasets/", "v003", "_r2025",
                 ".csv"),
          na = "")
```

# v085

## process data

```{r}
v085 <- get_sahie_measures(
  df_sahie = sahie_1, 
  df_fips = fips,
  age_cat = 0, # age: under 65
  vnum = "v085") %>% 
  add_flag_CT(vnum = "v085", old="U", new = "A") %>% 
  glimpse()

```

## save data

```{r}
write_csv(v085,
          paste0("../measure_datasets/", "v085", "_r2025",
                 ".csv"),
          na = "")
```

# v122

## process data

```{r}
v122 <- get_sahie_measures(
  df_sahie = sahie_1, 
  df_fips = fips,
  age_cat = 4, # age: under 19
  vnum = "v122") %>% 
  add_flag_CT(vnum = "v122", old="U", new = "A") %>% 
  glimpse()
```

## save data

```{r}
write_csv(v122,
          paste0("../measure_datasets/", "v122", "_r2025",
                 ".csv"),
          na = "")
```

