---
title: "NCHS v147_other Life expectancy (subgroups) - Rankings 2025, (use Census Pop Estimates + NCHS births for population ate groups)"
author: ganhua lu; modified by hannah olson-williams for rankings 2025
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(haven)
library(gt)

```

# about

## v147 other (subgroups)

-   2025 ranking

-   population datasets: 2020, 2021, 2022 

-   mortality datasets: 2020, 2021, 2022 

-   population sources: Census 5-year pop est + NCHS births (infant pop)

## age categories for LE calculations

```{r}
tibble::tribble(
  ~age_cat, ~`age`, 
   0L,        "0",       
   1L,      "1-4",       
   2L,      "5-9",       
   3L,    "10-14",       
   4L,    "15-19",       
   5L,    "20-24",       
   6L,    "25-29",       
   7L,    "30-34",       
   8L,    "35-39",       
   9L,    "40-44",       
  10L,    "45-49",       
  11L,    "50-54",       
  12L,    "55-59",       
  13L,    "60-64",       
  14L,    "65-69",       
  15L,    "70-74",       
  16L,    "75-79",       
  17L,    "80-84",       
  18L,      "85+",       
  ) 
```

## subgroups: 6 + 1

```{r}
tibble::tribble(
  ~subgroup, ~description, 
      
   1L,    "Non-Hispanic, White",       
   2L,    "Non-Hispanic, Black",       
   3L,    "Non-Hispanic, AIAN",       
   4L,    "Non-Hispanic, Asian",       
   5L,    "Non-Hispanic, NHOPI",       
   6L,    "Non-Hispanic, Two or more races (TOM)",   
  
   8L,    "Hispanic",  
  ) 
```

# ---------

# prepare

## set up directories for mortality data, pop data, raw data, processed data

```{r}
# mortality data, fixed-width txt files
nchs_mort_dir <- "E:/CHRR/mortality raw"
# census pop data, csv files
census_pop_dir <- "E:/CHRR/pop raw" 

# raw data directory
#data_raw_dir <- "../01_data/raw"
# director for processed data
#data_out_dir <- "../01_data/processed"


```

## load helper functions

-   some functions that can be used in general for mortality/natality measures

```{r}
source("nchs_measure_helpers.R")
```

## standard county and state fips

-   need for adding statecode in mortality data
-   need for constructing a dataset with starndard FIPS code for infant pop

```{r}

cfips = haven::read_sas("P:/CH-Ranking/Data/2025/2 Cleaned data ready for Calculation or Verification/county_fips_with_ct_old.sas7bdat")
sfips = haven::read_sas("P:/CH-Ranking/Data/2025/2 Cleaned data ready for Calculation or Verification/state_fips.sas7bdat")
fips = rbind(cfips, sfips)


```

# ---------

# pop data

## NCHS birth data: age_cat = 0, by subgroups

-   `nchs_births_2016_to_2022.csv`: prepared using NCHS natality data from 2016-2022

-   6 columns

    *   `race`: 1-White; 2-Black; 3-AIAN; 4-Asian; 5-NHOPI; 6-TOM
    
    *   `hisp`: 0-Not Hispanic; 1-Hispanic; 99-Unknown

```{r}
yrs <- c(2020:2022)

nchs_births <- read_csv(file.path("E:/CHRR/natality raw/nchs_births_2016_to_2022.csv"),
                        show_col_types = FALSE) %>% 
  filter(year %in% yrs) %>% 
  glimpse()

infant_pop <- nchs_births %>% 
  # exclude Hispanic unknown
  filter(hisp != 99) %>% 
  # set race/eth groups as 6 + 1
  mutate(race = case_when(
    hisp == 0 & race == 1 ~ 1, # non-Hispanic, "White",
    hisp == 0 & race == 2 ~ 2, # non-Hispanic, "Black",
    hisp == 0 & race == 3 ~ 3, # non-Hispanic, "AIAN", 
    hisp == 0 & race == 4 ~ 4, # non-Hispanic, "Asian",
    hisp == 0 & race == 5 ~ 5, # non-Hispanic, "NHOPI", 
    hisp == 0 & race == 6 ~ 6, # non-Hispanic, "Two or More Races"
    hisp == 1  ~  8,           # Hispanic
    TRUE ~ NA
  )) %>% 
  group_by(statecode, countycode, race) %>% 
  summarise(pop = sum(births, na.rm = TRUE), age_cat = 0, .groups = "drop") %>% 
  glimpse()
```

## Census pop: age_cat 1-18, by subgroups

### function to get pop by by race, age cat 1-18

```{r}
get_pop_by_sub61_agecat_v147 <- function(census_ct_estimates, year_num = 13, long = TRUE) {
  
  print(paste0("YEAR = ", year_num))
  
  pop_census_ct <- census_ct_estimates %>%  
    filter(year == year_num, agegrp %in% c(1:18)) %>% 
    # select columns for single races
    select(statecode = state, countycode = county, agegrp,
           # Non-Hispanic: 6
           starts_with("nhwa_"), # Not Hispanic, White alone
           starts_with("nhba_"), # Not Hispanic, Black or African American alone
           starts_with("nhia_"), # Not Hispanic, American Indian and Alaska Native alone
           starts_with("nhaa_"), # Not Hispanic, Asian alone
           starts_with("nhna_"), # Not Hispanic, Native Hawaiian and Other Pacific Islander alone
           starts_with("nhtom_"), # NHTOM_MALE	Not Hispanic, Two or More Races male population
           # Hispanic: 1
           starts_with("h_"), # Hispanic
    ) %>%  
    mutate(
      # 6: Non-Hispanic groups
      nh_white = nhwa_male  + nhwa_female,
      nh_black = nhba_male  + nhba_female,
      nh_aian  = nhia_male  + nhia_female,
      nh_asian = nhaa_male  + nhaa_female,
      nh_nhopi = nhna_male  + nhna_female,
      nh_tom   = nhtom_male + nhtom_female,
      # 1: Hispanic 
      hispanic = h_male  + h_female
    ) %>% 
    select(-c(nhwa_male:h_female)) %>% 
    # fix counties
    mutate(countycode = case_when(
      statecode=='46' & countycode =='113' ~ '102',
      statecode=='51' & countycode =='515' ~ '019',
      statecode=='02' & countycode =='270' ~ '158',
      statecode=='02' & countycode =='201' ~ '198',
      TRUE ~ countycode
    )) 

  # county pop: age cat 1-18; need to fix age_cat 1
  ct_pop_agecat_1_18 <- pop_census_ct %>%  
    filter(agegrp %in% c(1:18)) %>% 
    rename(age_cat = agegrp) %>% 
    group_by(statecode, countycode, age_cat) %>% 
    summarise(across(c(nh_white:hispanic), ~sum(., na.rm = TRUE)), .groups = "drop")

  # state pop
  st_pop_agecat_1_18 <- ct_pop_agecat_1_18 %>%
    mutate(countycode = "000") %>% 
    group_by(statecode, countycode, age_cat) %>% 
    summarise(across(c(nh_white:hispanic), ~sum(., na.rm = TRUE)), .groups = "drop")
  
  # us pop
  us_pop_agecat_1_18 <- ct_pop_agecat_1_18 %>% 
    mutate(statecode = "00", countycode = "000") %>% 
    group_by(statecode, countycode, age_cat) %>% 
    summarise(across(c(nh_white:hispanic), ~sum(., na.rm = TRUE)), .groups = "drop")
  
  ## county + state + us data
  pop_all_agecat_1_18 <- bind_rows(
    us_pop_agecat_1_18, 
    st_pop_agecat_1_18, 
    ct_pop_agecat_1_18) %>% 
    arrange(statecode, countycode, age_cat)
  
  # long form or wide form
  if(long){
    
    pop_all_agecat_1_18_long <- pop_all_agecat_1_18 %>% 
      pivot_longer(cols = c(nh_white:hispanic), names_to = "race", values_to = "pop") %>% 
      mutate(race = case_when(
        race == "nh_white" ~ 1,
        race == "nh_black" ~ 2,
        race == "nh_aian"  ~ 3,
        race == "nh_asian" ~ 4,
        race == "nh_nhopi" ~ 5,
        race == "nh_tom"   ~ 6,
        
        race == "hispanic"  ~ 8,
        TRUE ~ NA_real_))%>% 
      arrange(statecode, countycode, race, age_cat)
    
    return(pop_all_agecat_1_18_long)
    
  }
  else{
    return(pop_all_agecat_1_18)
  }
  
}
```

### function to process Census pop data for v147

```{r}
process_pop_v147 <- function(year) {
  pop <- read_census_pop(
    file_name = get_census_csv_file_info(year)$file_name
  ) %>%
    get_pop_by_sub61_agecat_v147(
      .,
      year_num = get_census_csv_file_info(year)$year_num
    )

  return(pop)
}
```


### 2022

```{r}
pop_2022 <- process_pop_v147(2022) %>% 
  glimpse()
```


### 2021
```{r}
pop_2021 <- process_pop_v147(2021) %>% 
  glimpse()
```

### 2020
```{r}
pop_2020 <- process_pop_v147(2020) %>% 
  glimpse()

```



## pop: age_cat 1-18 for 2019, 2020, 2021 by subgroups

```{r}
pop_agecat_1_18 <- bind_rows(  
  pop_2022,
  pop_2020,
  pop_2021
  ) %>%  
  group_by(statecode, countycode, age_cat, race) %>%  
  summarise(pop = round(sum(pop, na.rm = TRUE), digits = 0), .groups = "drop") %>% 
  # remove 02261
  filter(!(statecode == "02" & countycode =="261"  ) )

pop_agecat_1_18
```

## pop_all: add age_cat 0 (infant), fix age_cat 1

```{r}
# fix age_cat 1: subtract infant pop from NCHS births data
pop_agecat_1_18 %>% 
  filter(age_cat==1) %>% 
  bind_rows(infant_pop) %>% 
  arrange(statecode, countycode, age_cat, race) %>% 
  pivot_wider(names_from = age_cat, values_from = pop, names_glue = "pop_{age_cat}") %>% 
  mutate(pop = if_else(is.na(pop_0), pop_1, pop_1 - pop_0)) %>% 
  select(-c(pop_0, pop_1)) %>% 
  mutate(age_cat = 1)

```

```{r}
pop_all <- 
  bind_rows(
    # age_cat = 2 to 18
    pop_agecat_1_18 %>% filter(age_cat!=1),
    
    # age_cat = 0
    c(1:6, 8) %>% 
      map_df(~mutate(fips %>% select(1:2), race = .)) %>% 
      mutate(age_cat = 0) %>% 
      left_join(infant_pop, by = c('statecode', 'countycode', 'race', 'age_cat')) %>% 
      mutate(pop = if_else(is.na(pop), 0, pop)),
    
    # fixed age_cat = 1
    pop_agecat_1_18 %>% 
      filter(age_cat==1) %>% 
      bind_rows(infant_pop) %>% 
      pivot_wider(names_from = age_cat, values_from = pop, names_glue = "pop_{age_cat}") %>% 
      mutate(pop = if_else(is.na(pop_0), pop_1, pop_1 - pop_0)) %>% 
      select(-c(pop_0, pop_1)) %>% 
      mutate(age_cat = 1)
    ) %>% 
  arrange(statecode, countycode, age_cat, race) %>% 
  # set negative pop as 0; example: 15005
  mutate(pop = if_else(pop<0, 0, pop)) %>%
  # remove AK 02261
  filter(!(statecode == "02" & countycode == "261") ) %>% 
  glimpse()

pop_all  
```

```{r}
pop_all %>% 
  filter(pop < 0)

pop_all %>% 
  filter(statecode == "00", age_cat == 0) %>% 
  summarise(pop_tot = sum(pop))
```


# ---------

# mort data

-   Note: for convenience,  all mortality data files were named in the format as `MULTYEARUS.AllCnty.txt`, e.g., `MULT2019US.AllCnty.txt`

## header

```{r}
mort_header <- tribble(
  ~fld_name, ~start,  ~end,
  "state_of_residence", 29,  30,
  "county_of_residence", 35,  37,
  "detail_age", 70, 73,
  # "icd_code", 146, 149,
  "race", 445, 446,
  "race_recode_5", 450, 450,
  "hispanic_origin", 484, 486,
  "hispanic_origin_race_recode", 488, 488,
  "race_recode_40", 489, 490
)
```

## function to get age category for mort data
```{r}
get_age_cat_v147 <- function(age) {
  case_when(
    age == 0 ~ 0,
  	age<=4  ~ 1,
  	age<=9  ~ 2,
  	age<=14 ~ 3,
  	age<=19 ~ 4,
  	age<=24 ~ 5,
  	age<=29 ~ 6,
  	age<=34 ~ 7,
  	age<=39 ~ 8,
  	age<=44 ~ 9,
  	age<=49 ~ 10,
  	age<=54 ~ 11,
  	age<=59 ~ 12,
  	age<=64 ~ 13,
  	age<=69 ~ 14,
  	age<=74 ~ 15,
  	age<=79 ~ 16,
  	age<=84 ~ 17,
    age>=85 ~ 18,
    TRUE ~ NA
  )
}
```

## function to process single year mort data

```{r}
get_mort_by_sub61_agecat_v147 <- function(nchs_mort, df_fips){
  mort_1 <- nchs_mort %>% 
      # include only 50 states + DC
    filter(state_of_residence %in% c(state.abb, "DC")) %>% 
    rename(state = state_of_residence, countycode = county_of_residence) %>% 
    # fix county FIPS codes
    mutate(countycode = case_when(
      state =='SD' & countycode =='113' ~'102',
      state =='AK' & countycode =='270' ~'158',
      state =='VA' & countycode =='515' ~'019',
      state =='AK' & countycode =='063' ~'261',
      state =='AK' & countycode =='066' ~'261',
      TRUE ~ countycode
      )
    )
  
  mort_2 <- mort_1 %>% 
    # remove age not stated
    filter(detail_age!= 9999, detail_age!=1999) %>% 
    mutate(age_death = case_when(
      detail_age >1999 ~ 0,
      detail_age>1000 & detail_age<2000 ~ detail_age-1000,
      TRUE ~ NA_real_
    )) %>% 
    filter(!is.na(age_death)) %>% 
    select(-c("detail_age")) %>% 
    mutate(age_cat = get_age_cat_v147(age_death)) %>% 
    # race
    select(-race) %>% 
    mutate(race_recode_40 = as.numeric(race_recode_40)) %>% 
    mutate(
    race = case_when(
      # non-Hispanic
    	hispanic_origin<200 & race_recode_40==1 ~ 1, # nh_white
    	hispanic_origin<200 & race_recode_40==2 ~ 2, # nh_black
    	hispanic_origin<200 & race_recode_40==3 ~ 3, # nh_aian
    	hispanic_origin<200 & race_recode_40 %in% c(4:10)  ~ 4, # nh_Asian
    	hispanic_origin<200 & race_recode_40 %in% c(11:14) ~ 5, # nh_NHOPI
    	hispanic_origin<200 & race_recode_40 %in% c(15:40) ~ 6, # nh_two or more races
    	
      # Hispanic
  	  hispanic_origin<300  ~ 8, # Hispanic

    	TRUE ~ NA_real_
    )
  )
  
  # county
  mort_ct <- mort_2 %>% 
    group_by(state, countycode, age_cat, race) %>% 
    summarise(deaths = n(), .groups = "drop")
  
  # state
  mort_st <- mort_2 %>% 
    group_by(state, age_cat, race) %>% 
    summarise(countycode = "000", deaths = n(), .groups = "drop") 
  
  # us
  mort_us <- mort_2 %>% 
    group_by(age_cat, race) %>% 
    summarise(state = "US", countycode = "000", deaths = n(), .groups = "drop")
  
  # county + state + us
  mort_all <- bind_rows(mort_ct, mort_st, mort_us) %>% 
  arrange(state, countycode) %>% 
  # add statecode
  left_join(df_fips %>% select(state, statecode) %>% 
              distinct(),
            by = c("state")) %>% 
  ungroup() %>% 
  select(-state) %>% 
  select(statecode, everything()) %>% 
  arrange(statecode, countycode, race, age_cat) 
  
  return(mort_all)
  
}
```

## function to process mortality data for v147 subgroups

-   a helper that combines reading and processing one mortality dataset

```{r}
process_mort_v147 <- function(year){
  
  print(get_mort_file_name(year))
  
  mort <- read_mort_data(
    file_path = file.path(nchs_mort_dir, get_mort_file_name(year) ), 
    header = mort_header,
    n_max = Inf) %>% 
    get_mort_by_sub61_agecat_v147(df_fips=fips)
  
  return(mort)
}
```

## mort 2022

```{r}
mort_by_age_cat_2022 <- process_mort_v147(2022) %>% 
  glimpse()
```

## mort 2021

```{r}
mort_by_age_cat_2021 <- process_mort_v147(2021) %>% 
  glimpse()
```

## mort 2020

```{r}
mort_by_age_cat_2020 <- process_mort_v147(2020) %>% 
  glimpse()
```



## mort 3 years
```{r}
mort_all <- bind_rows(
  mort_by_age_cat_2020,
  mort_by_age_cat_2021, 
  mort_by_age_cat_2022
  ) %>% 
  group_by(statecode, countycode, age_cat, race) %>% 
  summarise(deaths = sum(deaths), .groups = "drop") %>% 
  glimpse()

```

# ----------

# Life Expectancy (LE) CALCULATION 

## merge mort and pop data

```{r}
merged <- pop_all %>% 
  left_join(mort_all %>% ungroup(), 
            by = c("statecode", "countycode", "age_cat", "race")) %>% 
  # replace NA with 0 for deaths
  mutate(deaths = if_else(is.na(deaths), 0, deaths)) %>% 
  glimpse()
  
```

## construct LE table

-   $x_i$ : Age at Start of Interval (Years)

-   age_interval: Age Interval (Years) 	

-   $n_i$ : Interval Width (Years)

-   $a_i$: Average proportion of age interval lived for those who die	

-   $deaths_i$ :Number Of Deaths In Interval 

-   $pop_i$ : Population Years At Risk 

-   $m_i$  : Death Rate In Age Interval

$$m_i = \frac{death_i}{pop_i}$$

-   $q_i$  : Probability Of Dying In Interval	

$$q_i = \frac{n_i\cdot{ m_i}}{1+(1-a_i)\cdot n_i\cdot m_i}$$

Ref: Chin Long Chiang, On Constructing Current Life Tables, Journal of the American Statistical Association, 1972


-   $p_i$  : Probability of Surviving the Interval

$$p_i = 1-q_i$$

### a helpter table

```{r}
helper_tbl <- tibble::tribble(
  ~age_cat, ~age_interval,  ~n_i,    ~a_i,  
   0L,        "0",        1L,   0.1,
   1L,      "1-4",        4L,   0.5,
   2L,      "5-9",        5L,   0.5,
   3L,    "10-14",        5L,   0.5,
   4L,    "15-19",        5L,   0.5,
   5L,    "20-24",        5L,   0.5,
   6L,    "25-29",        5L,   0.5,
   7L,    "30-34",        5L,   0.5,
   8L,    "35-39",        5L,   0.5,
   9L,    "40-44",        5L,   0.5,
  10L,    "45-49",        5L,   0.5,
  11L,    "50-54",        5L,   0.5,
  12L,    "55-59",        5L,   0.5,
  13L,    "60-64",        5L,   0.5,
  14L,    "65-69",        5L,   0.5,
  15L,    "70-74",        5L,   0.5,
  16L,    "75-79",        5L,   0.5,
  17L,    "80-84",        5L,   0.5,
  18L,      "85+",        NA,   0.5,
  )
```

### LE table

```{r}
le_table <- merged %>% 
  rename(deaths_i = deaths, pop_i = pop) %>% 
  left_join(helper_tbl, by = "age_cat") %>% 
  group_by(statecode, countycode, race) %>% 
  # calculate rates
  mutate(m_i = deaths_i/pop_i,
         q_i = n_i * m_i / (1 + (1 - a_i) * n_i * m_i),
         p_i = 1 - q_i
    ) %>%  
  mutate(
    q_i = if_else(age_cat == 18, 1, q_i), # fix for xi = 85
    p_i = if_else(age_cat == 18, 0, p_i)  # fix for xi = 85
  ) %>% 
  # Hypothetical Life Table Cohort
  mutate(l_i = accumulate(q_i, ~ .x * (1 - .y), .init = 100000)[-length(q_i)-1],
         d_i = q_i*l_i,
         L_i = (l_i-d_i)*n_i + (d_i * n_i * a_i))  %>%  
  mutate(L_i = if_else(age_cat == 18, l_i/m_i, L_i)) %>% # fix for xi = 85
  mutate(T_i = rev(cumsum(rev(L_i)))) %>%  # reverse cumsum
  mutate(e_i = T_i / l_i)

le_table
```

## calculate variance and confidence interval

-   Calculate var(qi) = (qi2*(1-qi))/Deaths, var(qi) = 0 where Deaths=0

    *   the (binomial) variance of the survival probability 𝑝𝑥 for the 𝑥𝑡ℎ age interval.

$When\space death_i\neq 0$

$$var(q_i)=\frac{q_i^2\cdot(1-q_i)}{death_i}$$

$When\space death_i=0$

$$var(q_i)=0$$

-   Calculate v1 = li2*(((1-ai)*ni+ei+1)2)*var(qi)

$$v1=l_i^2\cdot [(1-a_i)\cdot n_i + e_{i+1}]^2\cdot var(q_i)$$

-   Calculate v2 = v1 + v2i+1, v2w=0

$$v2_i = v1_i +v2_{i+1}$$
v2: Reverse cumulative sum of v1

-   Calculate var(ei) = v2/(li2)

$$var(ei)=\frac{v2}{l_i^2}$$

-   Calculate paw = lw/li

$$paw=\frac{l_w}{l_i}$$

-   Calculate term_adj = paw2/(Mw3*Populationw)

$$ term_{adj} = \frac{paw^2}{M_w^3\cdot Population_w^2}$$

-   Calculate var_adj = var(ei)+term_adj

$$var\_adj = var(e_i)+term_{adj}$$

-   Calculate standard error SE(ei) = sqrt(var_adj)

$$SE(e_i)=\sqrt {var\_adj}$$

-   Calculate CILL and CIUL = ei(+ or -)1.96*SE(ei) 

$$ CILL\space or\space CIUL=e_i \pm SE(e_i) $$

```{r}
le_table_var <- le_table %>% 
  group_by(statecode, countycode, race) %>%
  mutate(var_qi = case_when(
    deaths_i == 0 ~ 0,
    deaths_i != 0 ~ q_i^2 * (1 - q_i)/deaths_i
  )) %>% 
  mutate(e_i_lead = lead(e_i)) %>% #"next" (lead()) values in e_i; (e_i+1, or move one row up)
  mutate(v1 = l_i^2 * ((1-a_i)*n_i+e_i_lead)^2 * var_qi) %>% 
  select(-e_i_lead) %>% # remove column e_i_lead
  mutate(v2 = rev(cumsum(rev(replace_na(v1, 0))))) %>% 
  mutate(var_ei = v2 / l_i^2) %>% 
  mutate(paw = l_i[length(l_i)]/l_i) %>% 
  mutate(term_adj = paw^2 / (m_i[length(m_i)]^3 * pop_i[length(pop_i)])) %>% 
  mutate(var_adj = var_ei + term_adj) %>% 
  mutate(se = sqrt(var_adj),
         ci = se * 1.96,
         ci_lo = e_i - ci,
         ci_hi = e_i + ci,
         rse = se / e_i)

le_table_var
```

## LE from birth

```{r}
v147 <- le_table_var %>% 
  filter(age_cat == 0) %>% 
  select(statecode, countycode, race, rawvalue = e_i, cilow = ci_lo, cihigh = ci_hi, rse)

v147 
```

## suppression: pop < 5000 or RSE > 0.25

-   suppression for counties with pop < 5000
-   suppression for RSE > 0.25

```{r}
le_suppress <- le_table %>% 
  select(statecode, countycode, race, pop_i) %>% 
  group_by(statecode, countycode, race) %>% 
  summarise(pop = sum(pop_i, na.rm = TRUE), .groups = "drop") %>% 
  mutate(suppress = if_else(pop<5000, 1, 0)) 

```

```{r}
v147_2 <- v147 %>% 
  ungroup() %>% 
  left_join(le_suppress %>% select(-pop),
            by = c("statecode", "countycode", "race")) %>% 
  mutate(across(c('rawvalue', 'cilow', 'cihigh'), ~if_else(suppress == 1, NA, .))) %>% 
  select(-suppress) %>% 
  # replace inf with NA
  mutate(across(c('rawvalue', 'cilow', 'cihigh'), ~if_else(is.infinite(.), NA, .))) %>% 
  # suppression: RSE > 0.25
  mutate(across(c('rawvalue', 'cilow', 'cihigh'), ~if_else(rse > 0.25, NA, .))) %>% 
  select(-rse) %>% 
  glimpse()

```

## explore data

```{r}
summary(v147_2)

# counties
v147_2 %>% 
  filter(countycode != "000") %>% 
  select(-c(statecode, countycode,race)) %>% 
  skimr::skim()
```

### 6+1
```{r}
tab_1 <- print_summary_sub6_1(v147_2, 
                               'v147 Life expectancy (subgroups) R2025',
                               digits = 1)

tab_1

# tab_1  %>% 
#   gtsave("v147.png", expand = 10)


```

## get subgroup data in wide format

### function to get a subgroup

```{r}
get_race_data <- function(df, race_cat = 1, race_group = "v147_race_white"){
  v147_2 %>% 
    filter(race == race_cat) %>% 
    select(-race) %>% 
    rename({{race_group}} := rawvalue) %>% 
    rename_with(.fn = ~paste0(race_group, "_", .x), .cols = c("cilow" , "cihigh" ))
}
```

### subgroup data

```{r}
white <- get_race_data(v147_2, race_cat = 1, race_group = "v147_race_white")
black <- get_race_data(v147_2, race_cat = 2, race_group = "v147_race_black")
aian  <- get_race_data(v147_2, race_cat = 3, race_group = "v147_race_aian")
asian <- get_race_data(v147_2, race_cat = 4, race_group = "v147_race_asian")
nhopi <- get_race_data(v147_2, race_cat = 5, race_group = "v147_race_nhopi")
tom   <- get_race_data(v147_2, race_cat = 6, race_group = "v147_race_tom")

hispanic <- get_race_data(v147_2, race_cat = 8, race_group = "v147_race_hispanic")

```

### bind columns

```{r}
v147_other <- purrr::reduce(
  list(fips %>%  
         select(1:2), 
       white, black, aian, asian, nhopi, tom,
       hispanic
       ), 
  dplyr::left_join, 
  by = c("statecode", "countycode")) 

v147_other
```

# ---------

# save data

```{r eval=FALSE}
# long
write_csv(v147_2 %>% 
            rename_with(.fn = ~paste0("v147_", .x), .cols = c("rawvalue", "cilow" , "cihigh" )),
  file = "~/chrr_measure_calcs/duplicated_data/v147_other_r2024_long.csv")

# wide
write_csv(v147_other,
   file = "~/chrr_measure_calcs/duplicated_data/v147_other_r2024_wide.csv")
```

#----------
