---
title: "v181 - Library Access"
author: "how"
format: html
editor: visual
---

## Description: 
Library visits per person living within the library service area per year.

## Data Source: 
Institute of Museum and Library Services

Data Download Link: 
https://www.imls.gov/research-evaluation/surveys/public-libraries-survey-pls

FY2022 data was used for 2025 Rankings


## Measure Calculation: 

### National

National values are calculated like state values. The sum of library visits across the country and the national population are calculated, then the two numbers are divided.


### State-level
The state value is calculated by summing the library visits and the population from that state, then dividing the two sums.


### County-level 

Since a library's service area may cross multiple counties, we calculate the median number of visits for all libraries with service areas in the county and use this as the county value.

1. Bring in and clean library data files

Library data is provided in an outlet file (pls_fy22_outlet_pud22i) and an AE file (PLS_FY22_AE_pud22i).

The following code is provided by the data source:

Outlet file:

array num _numeric_;
do over num; * * if num = -1 then num = .M; /*recode missing value into .M*/
if num = -3 and STATSTRU ='23' then num = .C; /*recode Temporary Closed Library into .C*/
if num = -4 then num = .N; /*recode "Not Applicable" into .N*/
end; array char _character_;
do over char; if char ='M' then char = ' '; /*recode missing value into M for character variables*/
end; /*recode the rest of special missing into corresponding missing values*/

Apply the following criteria to clean the files:

Outlet file:

if STATSTRU in ('23') then delete; ***if they were temporarily closed during this year**;
if phone in ('-3') then delete;
if hours =-3 then delete;
if wks_open < 2 then delete; **if they were open less than 2 weeks**;

IF STATSTRU = '10' THEN DELETE; *this denotes an incorrect record in the data set;

if C_FSCS = "N" then delete; ***not a public library***;
if C_OUT_TY = "BM" then delete; ***if the outlet was a book by mail only***;
if STABR = "AS" or STABR = "GU" or STABR = "MP" or STABR = "PR" or STABR = "VI" then delete; **american samoa, guam, mariana islands, PR, virgin islands***;

AE file:

if STARTDAT = '-3' then STARTDAT = ' ';
if ENDDATE = '-3' then ENDDATE = ' ';
IF C_LEGBAS = 'SD' THEN DELETE; *delete school district libraries that wouldn’t be for the general public;

if C_FSCS = "N" then delete; **not a public library**;
if STABR = "AS" or STABR = "GU" or STABR = "MP" or STABR = "PR" or STABR = "VI" then delete; **american samoa, guam, mariana islands, PR, virgin islands***;
if STATSTRU in ('-23') then delete; ***if they were temporarily closed during this year**;
if visits IN (-3, -1, 0) then delete;
if phone in ('-3') then delete;
if HRS_open =-3 then delete;

NOTE: take a substring of CENTRACT to get state (SUBSTR(CENTRACT,1,2); and county (SUBSTR(CENTRACT,3,3) fips, but ensure you label the resulting variables to differentiate between the fips originating from the outlet file vs. the AE file to avoid overwriting during merge.

2. Calculate visits per population from AE file

visits_per_pop=visits/popu_und

3. Merge the Outlet and AE file by FSCSKEY 

Delete any rows where variable hours is missing--if hours =. then delete; 

4. Perform PROC UNIVARIATE (or similar) to calculate the median value of visits_per_pop and CIs

proc univariate DATA=[have] cipctldf alpha=0.05 noprint; 
var [variable name for visits/population];
by statecode_outlet countycode_outlet; *use the fips from the outlet file;
OUTPUT OUT=visit_median_ci pctlpts=50 pctlpre=p
cipctldf=(lowerpre=LCL upperpre=UCL);
run;

Any rows where cilow=cihigh should have their cilow and cihigh set to missing

5. Merge with county dataset so any missing counties can be included

6. Calculate state values from AE file

7. Calculate national values from AE file

8. Calculate state and national CIs 

Similar to how we generally calculate CIs.

9. Merge and clean up final dataset to follow CHRR formatting of measure datasets


# import packages

```{r}
library(tidyverse)
library(haven)
library(here)

```

# standard county and state fips

- The final dataset will need to be aligned with standard FIPS codes and to include all counties and states.
```{r}
fips <-  bind_rows(
  read_sas(here("inputs/county_fips_with_ct_old.sas7bdat")) ,
  read_sas(here("inputs/state_fips.sas7bdat")))  %>% 
  arrange(statecode, countycode) %>% 
  glimpse()
```

# import raw data

```{r}
aeraw = read.csv("raw_data/IMLS/PLS_FY22_AE_pud22i.csv")
outletraw = read.csv("raw_data/IMLS/pls_fy22_outlet_pud22i.csv")

```

# clean and calculate state and national values 

```{r}

aeclean = aeraw %>% 
  #filter(!is.na(HOURS)) %>% 
  filter(!(STATSTRU %in% c(23))) %>% #c(10, 23))) %>% 
  filter(!(PHONE %in% c(-3))) %>% #, -4))) %>% 
  #filter(HOURS != -3) %>% 
  #filter(STARTDAT != -3) %>% #start and end date shoudl be chekced - unclear that they need to be deleted..... 
  #filter(ENDDATE != -3) %>% 
  filter(C_LEGBAS != "SD") %>% 
  filter(HRS_OPEN != -3) %>% 
  #filter(WKS_OPEN >= 2) %>% 
  filter(C_FSCS != "N") %>% 
  #filter(C_OUT_TY != "BM") %>% 
  filter(!(STABR %in% c("AS", "GU", "MP", "PR", "VI"))) %>% 
  filter(!(VISITS %in% c(-3, -1, 0))) %>% 
  mutate(totcode = str_pad(CENTRACT, width = 11, pad = "0", side = "left"),
    statecode = substr(totcode, 1, 2), 
         countycode = substr(totcode, 3,5),
    visits_per_pop = VISITS/POPU_UND)




statevals = aeclean %>% group_by(statecode) %>% 
  summarize(numerator = sum(VISITS, na.rm = TRUE),
            denominator = sum(POPU_UND, na.rm = TRUE), 
            med_visit = numerator/denominator) %>% 
  mutate(countycode = "000")

ntlval = aeclean %>%
  summarize(numerator = sum(VISITS, na.rm = TRUE),
            denominator = sum(POPU_UND, na.rm = TRUE), 
            med_visit = numerator/denominator) %>% 
  mutate(statecode = "00", 
         countycode = "000")

```


# now clean the outlet file and merge 

```{r}

outletclean = outletraw %>% 
  filter(!is.na(HOURS)) %>% 
  filter(!(STATSTRU %in% c(10, 23))) %>% 
  filter(!(PHONE %in% c(-3))) %>%  # , -4))) %>% 
  filter(HOURS != -3) %>% 
  filter(WKS_OPEN >= 2) %>% 
  filter(C_FSCS != "N") %>% 
  filter(C_OUT_TY != "BM") %>% 
  filter(!(STABR %in% c("AS", "GU", "MP", "PR", "VI"))) %>% 
  mutate(totcode = str_pad(CENTRACT, width = 11, pad = "0", side = "left"),
         statecode = substr(totcode, 1, 2), 
         countycode = substr(totcode, 3,5),
         onelib = 1) %>% 
  group_by(FSCSKEY) %>%
  mutate(nlibrary = n()) %>%
  ungroup()

outletae = merge(outletclean, aeclean, by = "FSCSKEY", all.x = TRUE)
```

# calculate medians and merge with fipscodes 

```{r}
meds = outletae %>% group_by(statecode.x, countycode.x) %>%
  rename(statecode = statecode.x, countycode = countycode.x) %>% 
  summarize(med_visit = median(visits_per_pop, na.rm = TRUE))


cfips = fips %>% filter(countycode != "000")

oa_cfips = merge(cfips, meds, by = c("statecode", "countycode"), all.x = TRUE) %>% filter(countycode != "000")


oatot = bind_rows(oa_cfips, statevals, ntlval)
```

# rename some columns and save measure data 

```{r}
libsfinal = oatot %>% rename(v181_rawvalue = med_visit,
                             v181_numerator = numerator, 
                             v181_denominator = denominator)
write.csv(libsfinal, file = here("measure_datasets/v181_r2025.csv"))

```

